---
title: "project_report"
author: "Jack A Goldman"
date: "2024-02-23"
output: html_document
---

```{r setup, include=FALSE, results = 'hide'}
knitr::opts_chunk$set(echo = TRUE)
my_path <- c("./src/")             # set your path
source_files <- list.files(my_path, "*.R$")  # locate all .R files
purrr::map(paste0(my_path, source_files), source)
```

# The impact of the spatial legacies of spruce budworm defoliation on wildfire burn severity and post-fire recovery

## Background

Highlight forest complexity - 
highlight uncertainty around sbw-fix interactions


Wildfire behaviour is driven in part by the abundance, structure and connectivity of combustible fuels, therefore it is expected that spruce budworm defoliation will have strong bottom up effects on wildfire behaviour. However, current evidence on the impact that the spatial legacies of spruce budworm defoliation have on wildfire behaviour reports positive, negative and neutral effects. Limited evidence for the impact of SBW on wildfire behaviour may be result of the measure used to describe wildfire behaviour or how SBW defoliation was quantified. We attempt to tease apart the relationship between SBW defoliation and wildfire behaviour by 1) modelling burn severity and post-fire recovery and 2) creating a composite of SBW defoliation by combining different attributes used to describe defoliation into a synthetic index. To date, there are a lack of studies that have investigate the role of SBW defoliation on either wildfire burn severity or post-fire recovery. Likewise, to our knowledge no study have tried to model SBW defoliation using a composite variable that accounts for the different was to quantify SBW defoliation.

In addition, previous studies are concerned whether between fire variation in wildfire behaviour can be attributed to defoliation. However, isolation the effects of defoliation is difficult due to the amount of noise inherent to wildfire behaviour. Wildfire behaviour can be influenced by stand and landscape level drivers such as weather, climate, forest structure and topography. The influence of these drivers may impede the ability of previous studies to detect a signal between defoliation and burn severity. In an attempt to correct for this, we analyzed wildfire severity and forest recovery within fires that had areas that were defoliated, and areas that were not defoliated. This approach allows us to control for other drivers, and isolate the effects of defoliation while also allowing for between fire comparison. 


## Questions

The overarching question is: *how do the spatial legacies of spruce budworm defoliation influence wildfire burn severity and forest recovery in the boreal forest?*

Specifically we will address the following questions:

1.  How does spruce budworm defoliation influence burn severity?
    I.  Is there a difference in median severity between defoliated and non-defoliated fires?
    II. Is there a difference in severity extremes between defoliated and non-defoliated fires
    III. Is there a difference in variability in burn severity between defoliated and non-defoliated fires
2.  How does spruce budworm defoliation influence post-fire recovery?
    I.  Is there a difference in forest recovery 10 years post-fire between defoliated and non-defoliated fires?
    II.  Is there a difference in forest recovery between 1-5 years and 6-10 years between defoliated and non-defoliated fires?
3.  Is the strength of the relationship between defoliation intensity (synthetic index) and burn severity or forest recovery greater than independent metrics of defoliation?
4.  Does defoliation intensity (synthetic index) capture more variation in burn severity and forest recovery relative to defoliation metrics modeled independently?

## Methods

This study focuses on wildfire that burned in the boreal shield ecoregion of Ontario. Our analysis was restricted to wildfires that burned between 1986 and 2013. 1985 was chosen as the first fire year in our dataset because it is the first year that remotely sensed burn severity data are available. Similarly, our analysis of forest recovery required landsat imagery for every year for 10 years post-fire, therefore 2013 is the last year where we were able to calculate remotely sensed imagery. We identified wildfires that were defoliated up to 15 years prior to the year of fire. We chose to limit the defoliation to within 15 years because previous studies have shown that beyond 15 years, defoliation is not significant due to enhanced regrowth. We only selected fires that contained both defoliated and non-defoliated areas. Once identified, we split each fire into two distinct areas. This resulted in a total of 34 (n = 68) fires containing both defoliated and non-defoliated areas.

Remotely sensed burn severity data was retrieved using Google Earth Engine and Landsat. To retrieve remotely sensed burn severity data for each fire, I will separate each individual fire event into the defoliation affected area and the unaffected area and retrieve burn severity maps for each perimeter. Burn severity data will be generated at the pixel level using Landsat data at 30m spatial resolution. I will then extract the burn severity value for each pixel and assign it to either the defoliated or non-defoliated category.

I will quantify time since defoliation as the difference between fire year and the last year a cell was defoliated. I will also include fire seasonality (Spring, Summer, Fall). Using the centroid of each 30m pixel I will interpolate drought code for the day that the fire burned. This will allow me to control for trends in weather at the time of the fire. Using Daymet V4, I will retrieve climate averages for the years between defoliation and fire in Google Earth Engine. The exact climate metrics are yet to be determined. I will gather elevation data for each pixel as well as slope and aspect in google earth engine using the Canadian Digital Elevation Model provided by NRCAN.

In the first part of our analysis we tested multiple hypotheses concerned with how burn severity or forest recovery differs following fires that burned in defoliated vs. non defoliated areas. To determine defoliated and non-defoliated areas, we identified areas where defoliation events overlapped wildfire perimeters and took the difference between the

### Structural Equation Model

The second part of the analysis looks the effects of defoliation intensity on burn severity and forest recovery following fire. Defoliation alters fuel structure and fuel load and is expected to cause influence fire behaviour. Previous studies in Ontario have shown that defoliation can alter fire behaviour for up to 15 following defoliation [@james2017], this period is commonly referred to as the "window of opportunity".

Spruce budworm defoliation is multivariate in nature. The effects of spruce budworm defoliation can be described by the presence and absence of defoliation, the time since defoliation to the time of fire and the number of years that an area was defoliated (a proxy for defoliation load). It is common practice for studies that concerned with the effects of spruce budworm defoliation on wildfire to use a since predictor, or build multiple sequential models to incorporate the multivariate nature of defoliation [@james2017 @james2011 @candau2018].

Structural equation models allow for the multivariate nature of defoliation by creating composite variable that allows us to combine the variation explained by multiple predictors into a synthetic index. Using structural equation models, we will combine create a synthetic index called defoliation intensity which combines defoliation presence/absence, time since defoliation and number of years defoliated.

We will fit these models to data of consisting of 34 paired defoliated/non-defoliated fires that burned in Ontario between 1986-2012. We will then model the causal effects of defoliation intensity on burn severity and forest recovery following fire. Forest recovery is measured by taking the mean composite of the normalized burn ratio for a given summer for 10 years for following fire. We then use a Theil-Sen regression to calculate the sen's slope across the 10 years which we call the slope of recovery or forest recovery.

In our model, defoliation intensity will effect for burn severity and forest recovery. Below is the directed acyclic graph (DAG) of our a priori causal hypothesis.

```{r, echo=FALSE}
DiagrammeR::grViz("Figures/chapter3-step1-dag.gv")


```


In this DAG we see that time since defoliation, presence/absence of defoliation (defoliated) and number of years defoliated are combined to form the synthetic index (defoliation intensity) which is then used to predict the effects on burn severity and forest recovery. Defoliation intensity will directly effect forest recovery by altering stand structure and forest composition prior to fire, which will impact the number and distribution species and their seed banks that are able to establish post fire, thus affecting the recovery trajectory. Similarly, defoliation intensity will effect burn severity by either increasing or decreasing severity which will lead to a change in the recovery trajectory by either increasing or decreasing the slope of recovery.

## Expected results



## Preliminary results

How does spruce budworm defoliation influence burn severity?

```{r, echo = FALSE}
severity_table(targets::tar_read(sev_ttest))
```

How does spruce budworm defoliation influence post-fire recovery?

```{r, echo = FALSE}
slope_table(targets::tar_read(slo_ttest))
```

Severity visualization
```{r, echo = FALSE}
targets::tar_read(lolliplot_sev)

```


#### Is there a difference in the direction of the effects ?



Hypothesis 4: 

## Significance

Land managers in the boreal forest region require information about burn severity to address immediate hazards and outcomes resulting from wildfires. Understanding how previous disturbance history shapes patterns of burn severity is needed to guide management activities and ensure downstream ecosystem provisioning when threatened by multiple disturbances. Results from this chapter will provide a framework for modelling the effects of spruce budworm defoliation on wildfire severity and help managers identify areas that are at the highest risk of losing valuable ecosystem services. 
